---
title: "Antibiotic Resistance"
author: "kammer"
date: "2024-11-29"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(DBI)
library(RSQLite)
library(viridis)
library(GGally)
library(lattice)
library(gt)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Antibiotic resistance is increasing world wide in pathogenic microbial populations. Therapeutic and prophylactic use of antibiotics constitutes selective pressure that favors the evolution of antibiotic resistance strains. The durability of a resistance mutation in pathogens is partly determined by it overall fitness to the organism. Resistant mutations usually target key biochemical processes and can potentially impart a fitness cost to the microorganism. The goal of this data investigation is to study how resistance gene mutations affect the overall fitness of a microorganism.

# Methods
Evaluation of the relationship between resistance mutations and fitness was by analyzing data from the article *The fitness costs of antibiotic resistance mutations* authored by *Melnyk, Wong and Kassen*.  The data focuses on fitness effects of single chromosomal mutations events that lead to the development of antibiotic resistance.  A special emphasis is placed on gram positive organisms ,specifically Staph aureus.



## Evaluate the ranges of numeric variables in the data set
```{r echo=FALSE, message=FALSE, warning=FALSE}
costR <-  read.csv("Raw_data/costR.csv", stringsAsFactors = FALSE)
costR %>% 
  select(fitness) %>% 
  ggplot(mapping = aes(x = fitness))+
  geom_histogram(binwidth = 0.01)+
  labs(title = "Range of Fitness Values")
  theme()

ggplot(data = costR, mapping = aes(x = SE))+
  geom_histogram(binwidth = 0.01)+
  labs(title= "Range of SE values")
  theme_classic()

ggplot(data = costR, mapping = aes(x = Selection_coefficient))+
  geom_histogram(binwidth = 0.01)+
  labs(title="Range of Selection coefficient values",
       x = "Selection Coefficient")
  theme_bw()

costR %>% 
  select(MIC) %>% 
  ggplot(mapping = aes(x = MIC))+
  geom_histogram(bins = 15)+
  labs(title = "Range of Minimum Inhibitory Concentration values")
  scale_fill_viridis_d(option = "magma")

```

# Relationship between Numeric Variables
The relationship between the numeric variables of 1)fitness, 2)SE and
3)Selection_coefficient are studied using 4 different approaches. 

Scatterplot with color or size Encoding. The x and y axis encodes two ot the three variables.  Color or size is used to represent the third variable.

```{r}
ggplot(data = costR, mapping = aes(x = fitness, y = Selection_coefficient, color = SE))+
  geom_point()+labs(title = "Scatter Plot with color encoding",
                    x="Fitness",
                    y = "Selection Coefficient",
                    color = "SE")

ggplot(data = costR, mapping = aes(x = fitness, y = Selection_coefficient, size = SE))+
  geom_point()+
  labs(title = "Scatter plot with size encoding",
       x = "Fitness",
       y = "Selection Coefficient",
       size = "SE")
```

## Pair Plot (Scatterplot Matrix). The GGally package is used to create a scatterplot matrix to visualize pairwise relationships between multiple variables.

```{r}
numericonly <- costR %>% 
  select(fitness, Selection_coefficient, SE)
#Use the data
ggpairs(numericonly, title = 'Pair Plot')

```

## Scatterplot Matrix using the Lattice package

```{r}
splom(numericonly,
      main = "Scatterplot Matrix")
```


# Which mutations increae or decrease fitness
```{r}

  costR %>% 
    select(gene_mutation, fitness) %>% 
    na.omit() %>% 
    filter(fitness <1.0) %>% 
    distinct(gene_mutation)
  

costR %>% 
    select(gene_mutation, fitness) %>% 
    na.omit() %>% 
    filter(fitness> 1.0) %>% 
    distinct()

```
No apparent relationship between increase or decrease in fitness and bacterial type

create ranges of fitness value and assign to general categories
```{r echo=FALSE}
costR %>% 
  mutate(GroupFIT = case_when(fitness > 1.35 ~ "Most Fit of All",
                              between(fitness, 1.25, 1.35)~ "Highly Fit",
                              between(fitness, 1.0, 1.25)~ "Fit",
                              between(fitness, .75, 1.0) ~ "Less Fit",
                              between(fitness, .35, .75)~ "Much Less Fit",
                              fitness<.35 ~ "Least Fit of All",
                              TRUE ~ "Unknown")) %>% 
  select(GroupFIT, bacteria) %>% 
  group_by(GroupFIT) %>% 
  tally() %>% 
  mutate(pct = round(100 *(n/sum(n)), digits =2)) %>% 
  arrange(factor(GroupFIT, levels = c("Most Fit of All", "Highly Fit",
                                      "Fit", "Less Fit",
                                      "Much Less Fit"),
                 "Least Fit of All")) %>% 
  gt() %>% 
  tab_header(title = "Fitness Groups") %>% 
  cols_label(GroupFIT = "Group",
             n = "Number of organisms",
             pct = "Percentage"
             )
```





The cell wall is an important target of mutations that
confer antibiotic resistance, and thus intuitively, it seems
possible that there may be a difference in fitness costs associated
with antibiotic resistance between Gram-positive,
which has a much thicker layer of peptidoglycan in their
cell wall, and Gram-negative bacteria, which has a much
thinner cell wall.

```{r}
GramFIT <- costR %>% 
  filter(gram_stain == "negative" | gram_stain == "positive") %>%
  group_by(gram_stain) %>% 
  summarise(mean_fitness = mean(fitness),
            IQR = IQR(fitness),
            Minumum = min(fitness),
            Maximum = max(fitness),
            Standard_deviation = sd(fitness))

GramFIT %>% 
  gt() %>% 
  tab_header(title = "Differences in fitness based on gram stain") %>% 
  cols_label(gram_stain = "Gram stain",
             mean_fitness = "Mean Fitness",
             IQR = "Interquantile Range",
             Minumum="Minimum",
             Maximum= "Maximum",
             Standard_deviation= "Standard Deviation")


```
Gram-positive bacteria had a significantly
greater fitness costs associated with resistance mutations
(mean fitness = 0.822) when compared with Gram-negative
bacteria (mean fitness = 0.973). Caution must be used in interpreting
this result because there were a greater number of Gram-negative
than Gram-positive bacteria in the data set.



Mean relative fitness and 95% CI of antibiotic Resistance mutation
associated with a specific antibiotic, Fitness cost < 1 indicates less fit

```{r}
costR %>% 
  select(antibioticType, fitness) %>% 
  group_by(antibioticType) %>% 
  summarise(mean_cost = mean(fitness),
            stddev = sd(fitness, na.rm = TRUE),
            meansd_l = mean_cost - stddev,
            meansd_u = mean_cost + stddev) %>% 
  ggplot(mapping = aes(x= mean_cost, y = antibioticType))+
  geom_point()+
  geom_vline(xintercept = 1)+
  geom_errorbar(mapping = aes(xmin = meansd_l, xmax = meansd_u))+
  labs(x = "Mean Fitness Cost", y = "Antibiotic")+
  ggtitle("Fitness Cost each Antibiotic")

```

Additionally, resistance mutations in two species, Enterococcus faecium and Borrelia burgorferi, showed no evidence for a cost of resistance on
average while those recovered from all other species were
on average costly. Mean relative fitness and 95% CI of antibiotic resistance mutation associated with a given bacteria Fitness cost < 1 indicates less fit.

```{r}
costR %>% 
  select(bacteria, fitness) %>% 
  group_by(bacteria) %>% 
  summarise(mean_cost = mean(fitness),
            stddev = sd(fitness, na.rm = TRUE),
            meansd_l = mean_cost - stddev,
            meansd_u = mean_cost + stddev) %>% 
  ggplot(mapping = aes(x = mean_cost, y =bacteria))+
  geom_point()+
  geom_vline(xintercept = 1)+
  geom_errorbar(mapping = aes(xmin = meansd_l, xmax = meansd_u))+
  labs(x = "Mean Fitness Cost", y = "Bacteria")+
  ggtitle("Fitness Cost per Bacterial Species")
```



create a heat map with antibiotic and bacteria as the two categorical variables
and fitness as a numeric variable.

```{r}
summary_data <- costR %>%
  drop_na() %>% 
  group_by(bacteria, antibioticType, fitness) %>%
  summarize(MeanFITNESS = mean(fitness))
ggplot(summary_data, aes(x = antibioticType, y = bacteria, fill = fitness)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Heat Map of Values by Antibiotic and Bacteria",
       x = "Antibiotic",
       y = "Bacteria",
       fill = "fitness")
```


## Analysis of Staph aureus gene mutations
What Staph aureus gene mutations associated with fitness is < 1, indicating less fit?

```{r}
BacteriaMoreFIT <- costR %>% 
  select(bacteria, gene_mutation, gene_target, fitness) %>% 
  filter(bacteria == "S. aureus" & fitness <1.0) %>% 
  distinct(gene_mutation)
```

What Staph aureus gene mutations associated with fitness is > 1, indicating improved fitness?

```{r}
BacteriaLessFit <- costR %>% 
  select(bacteria, gene_mutation, gene_target, fitness) %>% 
  filter(bacteria == "S. aureus" & fitness > 1.0) %>% 
  distinct(gene_mutation)
```

There are 5 gene mutations in Staph aureus associated with improved microbial fitness and 44 gene mutations associated with a decrease in fitness.