---
title: "practice"
author: "kammer"
date: "2024-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(knitr)
```


# Introduction

YAML is Yet Another Markdown Language.  You can do *italics* or you can do **bold** and include a [link](https://blog.djnavarro.net/posts/2023-04-12_metropolis-hastings/)


# R code

```{r our first chunk, eval=FALSE}
values <-  rnorm(9)
values
```


We can also use echo option to show or not show the code and the results to show or not show the output.


# More text tools

# subheadings

Unordered Lists:

-first item

-second item


Ordered Lists

1. first

2. second


We can insert an image like this: ![alt text](databases_WLF/Assignment_4_Kammer.pdf){width="20%}


We can *add* citations^[footnotes]

We can use **math mode** $\alpha \ beta$ and so on.


Make tables

```{r Penguins data set}
penguins_sm <-  head(penguins)
kable(penguins_sm)
```

$$
\begin{aligned}
\text{resemblance}_{ij} &= \hat{\gamma}_{0j} + 
\hat{\gamma}_1\text{Intervention}_{ij} + e_{ij}\\
\hat{\gamma}_{0} &= \hat{\gamma}_0 + \hat{u}_{0j}
\end{aligned}
$$
$$
\begin{aligned}
E[X] = \int_{-\infty}^{\infty} xf(x) dx; \\

&\int_{0}^{1} x dx = \frac12 x^2 |_0^1 = \frac12
\end{aligned}
$$



